CL-PFU: A suite of R packages for energy conversion chain analysis

### 2023-08-17 paper.Rmd
Add a message to editors.

<!-- Max length: 250--1000 words. At the moment, we are at 1444 words. -->
<!-- But I saw many JOSS papers are longer than 1000 words.  -->
<!-- And our paper is describing many packages, whereas -->
<!-- most JOSS papers describe only a single package. -->
<!-- So I think we are going to be OK. -->
<!-- I think we can own the issue in the cover letter,  -->
<!-- saying that we are willing to consider splitting into -->
<!-- two papers upon the editor's advice.  -->

@@

Energy flowing through societal energy conversion chains (ECCs)
enables economic activity and facilitates human flourishing.
To understand economic growth and human well-being, 
the field of societal energy analysis (among other activities)
evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, natural gas, wind, and solar) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal _exergy_ analysis (SEA), 
an extension of societal _energy_ analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.
See any number of web references for an exergy primer,
including @SciDirect:2023. 
**** Do we have a better primer? ****].

We created a suite of `R` packages 
to assist SEA practitioners to analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the CL-PFU database,^["CL-PFU" 
is an initialism for "Country-Level
Primary, Final, and Useful."
The CL-PFU database contains many countries as well as 
continental and world aggregations.
Technically speaking, 
we create [matsindf][link-matsindf] data frames
stored as `[pins][link-pins]` on a pinboard, 
not SQL or similar databases.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

The new `R` packages described in this paper
enable the creation of databases of primary, final, and useful
energy and exergy conversion chains.
Such databases will enable advances in the societal exergy analysis community
and lead to greater insights 
about the role of energy 
in economic growth and human well-being.
An example of early work using the CL-PFU database
is the calculation of useful stage 
Energy Return On Investment for fossil fuels [@Aramendia:2023aa], 
for which the [ECCTools][link-ecctools] [@Aramendia:2022aa] and
[EROITools][link-eroitools] [@Aramendia:2022ab] packages were developed, 
expanding the software capabilities introduced here.^[Note
that the [ECCTools][link-ecctools] and [EROITools][link-eroitools] packages were
not reviewed for this paper.]

<!-- # Mathematics -->

<!-- Single dollars ($) are required for inline mathematics e.g. $f(x) = e^{\pi/x}$ -->

<!-- Double dollars make self-standing equations: -->

<!-- $$\Theta(x) = \left\{\begin{array}{l} -->
<!-- 0\textrm{ if } x < 0\cr -->
<!-- 1\textrm{ else} -->
<!-- \end{array}\right.$$ -->

<!-- You can also use plain \LaTeX for equations -->
<!-- \begin{equation}\label{eq:fourier} -->
<!-- \hat{f}(\omega) = \int_{-\infty}^{\infty} f(x) e^{i\omega x} dx -->
<!-- \end{equation} -->
<!-- and refer to \autoref{eq:fourier} from text. -->

<!-- # Figures -->

<!-- Figures can be included like this: -->

<!-- ![Caption for example figure.\label{fig:example}](figure.png) -->
<!-- and referenced from text using \autoref{fig:example}. -->

<!-- Figure sizes can be customized by adding an optional second parameter: -->

<!-- ![Caption for example figure.](figure.png){ width=20% } -->

>>>>

<!-- Message to editors: -->

<!-- This paper describes a suite of packages -->
<!-- to assist SEA practitioners to analyze  -->
<!-- energy movement through society.  -->
<!-- The new packages enable analysis of any country in the world -->
<!-- across timespans of decades or longer. -->
<!-- In short, the packages enable, for the first time,  -->
<!-- scalable Societal Exergy Analysis (SEA). -->

<!-- In an email exchange in late July 2021,  -->
<!-- editor Afron Smith recommended that we  -->
<!-- create a separate repository for the paper,  -->
<!-- mentioning all of the individual packages in the manuscript. -->
<!-- Thus, the paper is slightly longer than 1000 words. -->
<!-- That said, we are willing to consider splitting into -->
<!-- two papers upon the editor's advice. -->

<!-- No portion of this paper has been submitted or published -->
<!-- in another peer-reviewed journal.  -->
<!-- A paper with a very different focus  -->
<!-- (efficiency results from the full CL-PFU database) -->
<!-- is in preparation. -->
<!-- Our desire is that the results paper will reference -->
<!-- this JOSS paper. -->

@@

Energy flowing through societal energy conversion chains (ECCs)
enables economic activity and facilitates human flourishing.
To understand economic growth and human well-being, 
the field of societal energy analysis (among other activities)
evaluates ECCs
from the primary stage 
(resources extracted from the environment, 
such as coal, oil, natural gas, wind, and solar) 
to the final stage
(energy purchased by consumers, 
such as refined petroleum and electricity),
to the useful stage
(energy desired by the end user, 
such as heat, motion, and light),
and sometimes to energy services
(such as thermal comfort, transport, and illumination).
Societal _exergy_ analysis (SEA), 
an extension of societal _energy_ analysis,
quantifies ECCs in exergy terms^[Exergy is 
the mechanical work potential of energy.
See any number of web references for an exergy primer,
including @SciDirect:2023.].

We created a suite of `R` packages 
to assist SEA practitioners to analyze 
energy movement through society. 
The new packages enable analysis of any country in the world
across timespans of decades or longer.
In short, the packages enable, for the first time, scalable SEA.
We used the new packages to create the CL-PFU database,^["CL-PFU" 
is an initialism for "Country-Level
Primary, Final, and Useful."
The CL-PFU database contains many countries as well as 
continental and world aggregations.
Technically speaking, 
we create [matsindf][link-matsindf] data frames
stored as [pins][link-pins] on a pinboard, 
not SQL or similar databases.] 
a new resource for the SEA community [@Marshall:2023aa].

@@

The new `R` packages described in this paper
enable the creation of databases of primary, final, and useful
energy and exergy conversion chains.
Such databases will enable advances in the societal exergy analysis community
and lead to greater insights 
about the role of energy 
in economic growth and human well-being.
An example of early work using the CL-PFU database
is the calculation of useful stage 
Energy Return On Investment for fossil fuels [@Aramendia:2023aa], 
for which the [ECCTools][link-ecctools] [@Aramendia:2022aa] and
[EROITools][link-eroitools] [@Aramendia:2022ab] packages were developed, 
expanding the software capabilities introduced here.^[Note
that the [ECCTools][link-ecctools] and [EROITools][link-eroitools] packages 
are not the subject of this paper.]
